---
output:
  github_document:
    toc: true
#always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R package: `algebraic.dist`

<!-- badges: start -->
<!-- badges: end -->

An algebra over distributions (random elements).

## Installation

You can install the development version of `algebraic.dist` from [GitHub](https://github.com/) with:

```{r,eval=F}
# install.packages("devtools")
devtools::install_github("queelius/algebraic.dist")
```
   
## TODO

1. Flesh out the vignettes to show how to use the package.

2. Reincorporate the `edist` (expresion distribution), which is just a
an R expression combined with an environment that has a dictionary for
each of the variables in the expression where if a variable is a `dist`-type
object, then we treat that variable like a random variable. When we, say,
sample from the `edist` object, we'll sample from the `dist`-type objects
in the environment and then evaluate the expression with the sampled values.

3. For taking the PDF, CDF, expectation, and so on, for now we just use the
sampler and then put it into a `empirical_dist` object, which permits those
kinds of operations, but there are clearly a lot of interesting things we can
do, e.g., `min(x1, x2)` when `x1` and `x2` are exponential distributions
is also an exponential distribution with a failure rate `lambda1 + lambda2`.
There are a host of relations between various named distributions, and there
are even more cases where the distribution, no matter how complex, has an
asymptotically normal distribution.