---
title: "Exponential distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exponential distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
To demonstrate the R package `algebraic.dst`, we consider the relatively
simple case of exponentially distributed random variables.

First, we load the R package `algebraic.dist` with:

```{r setup}
library(algebraic.dist)
```

# Generating a sample

We define the parameters of the i.i.d. random sample with:
```{r}
n <- 1000
rate <- 1
```

We generate a random sample $X_i \sim \operatorname{EXP}(\lambda=`r rate`)$ for
$i=1,\ldots,n$ with:
```{r,message=F,warning=F}
library(stats)
library(tidyverse)

exp.dist <- make_exp_dist(rate=1)
#x <- tibble(x=sample(exp.dist)(n))
```

We have observed a sample of size $n=`r n`$.
We show some observations from this sample (data frame) with:
```{r}
print(head(x,n=4))
```

We show a histogram of the sample overlaid with a plot of the exponential
function's pdf, with:
```{r,message=F,warning=F}
library(ggplot2)
ggplot(x, aes(x=x)) + geom_histogram(aes(y=..density..),alpha=.2) +
    xlim(0,6) + 
    geom_function(fun=dexp)
```

```{r}
exp.dist.add <- make_exp_dist(rate=3) + make_exp_dist(rate=2)
```
